\section[Fixed point theory for cpos]%
{Fixed point theory for complete partial orders}\label{sec-1}

% Further explanations: Winskel, p.72.
Before getting our hands dirty with category theory, it would be appropriate to
introduce the well-established framework for the solution of recursive domain
equations based on complete partial orders.
These structures, along with continuous functions over them, allow us to properly treat the foundational aspects of denotational semantics.

\subsection[Complete partial orders]{Complete partial orders}

A \emph{partially ordered} set, or \emph{poset}, consists of a set endowed with a reflexive, antisymmetric and transitive binary relation.
Typical examples of posets are:
\begin{itemize}
  \item the numerical sets \(\mathbb{N}\), \(\mathbb{Z}\), \(\mathbb{Q}\), \(\mathbb{R}\), endowed with the standard order relation;
  %\item numerical sets, endowed with the standard order relation;
  \item given a set \(X\), the power set \(2^X\) with the inclusion relation.
\end{itemize}
In domain theory, the elements of a poset are thought of as bits of information or results of a computation, partially ordered by the amount of information they convey, i.e.\ sorted out by an \emph{information ordering}, denoted by ``\(\sqsubseteq\)''. %chktex 21
% In domain theory, we can visualize the elements of a partial order as points of information; thus, we can consider the \emph{information ordering}, denoted with the symbol ``\(\sqsubseteq\)'', which sorts out the points of a poset by the amount of information they convey.

Taking the investigation further, it would also be desirable to be able to represent sequences where information consistently increases, that is, where each new element carries more information than the previous one.
% Taking the investigation further, it would also be desirable to model steps of consequent computations, where each new object carries more information than the previous one.
In a partial order \((D, {}\sqsubseteq)\), this can be represented with an
\emph{ascending \(\omega\)--chain}, which is an increasing sequence %chktex 21
\(d_0 \sqsubseteq d_1 \sqsubseteq \dotsb \sqsubseteq d_n \sqsubseteq \dotsb{}\) of elements of \(D\).

Following the analogy between semantical domains and mathematical objects, the \emph{least upper bound} of such an \(\omega\)--chain, if it exists, would be that element that combines anything but the information contained in the elements of the \(\omega\)--chain. %chktex 21

Going back to the example shown in the Introduction, the sequence
\begin{equation*}
  \emptyset, \lbrace (0,1) \rbrace, \lbrace (0,1), (1,1) \rbrace, \dotsc
\end{equation*}
is an \(\omega\)--chain in \(\mathbb{N}^\mathbb{N}\) and its least upper bound is the factorial function \(f(n) = n!\). %chktex 21

Those requirements, essential to succeed in the modelling of meaning domains, lead to the following definition.

\begin{dfn}
  A \emph{complete partial order}, or \emph{\(\omega\)--cpo}, is a partial order in which every ascending \(\omega\)--chain has a least upper bound. %chktex 21
  An \emph{\(\omega\)--cpo with bottom} is a complete partial order with a least element %chktex 21
  \(\bot\), called the \emph{bottom}. %chktex 21
\end{dfn}

\subsection[Continuous functions]{Continuous functions}

The first criterion to be respected in the construction of a theoretical model for the description of programs written in a certain programming language is that
\begin{center}
  The more input is provided, the more \((\ge)\) output is expected.
\end{center}
How can this be obtained? In mathematical terms, the notion of \emph{monotonic} functions is introduced, namely functions of posets%
\(f \colon (D_1,{}\sqsubseteq_1) \to (D_2,{}\sqsubseteq_2)\) such that
\(
  \forall{}x,y \in D_1.\; x \sqsubseteq_1 y \implies f(x) \sqsubseteq_2 f(y).
\)

Moreover, it is important that programs do behave in a (somehow) predictable way: one should reasonably expect that a computable function is one that behaves well with respect to least upper bounds
%going on with the computations, one would like to come to a conclusion.
This justifies the following definition.

\begin{dfn}
  A function of cpos \(f \colon D_1 \to D_2\) is \emph{continuous} whenever
  \begin{equation*}
    \bigsqcup_{n \in \mathbb{N}} f(d_n) = f\left(\bigsqcup_{n\in\mathbb{N}}d_n\right)
  \end{equation*}
  hold for each ascending \(\omega\)--chain \(\lbrace d_n \rbrace_{n \in \mathbb{N}} \subseteq D_1\). %chktex 21
\end{dfn}

\subsection[Fixed points]{Least fixed points and recursion}

Now that we've defined which structures underlie our mathematical framework, it is time to see how a formal foundation of denotational semantics is given, supporting both recursive programs and recursive data types, defined as least
fixed points of continuous functions over a complete partial order.
To back this claim up, we are going to state some fundamental results in the field.

\begin{lem}[least (pre)fixed point lemma]\label{lem:cpo-prefixed}
  Let \(f\) be a monotonic function from a poset \(D\) to itself.
  Then, if it exists, the least prefixed point of \(f\) is also its least fixed point.
\end{lem}
\begin{proof}
  First, if we assume that \(x\) is the least prefixed point of \(f\), then \(f(x)\) is also a prefixed point of \(f\), since \(f\) is monotonic:
  \(
    f(x) \sqsubseteq x \implies f(f(x)) \sqsubseteq f(x).
  \)
  
  Moreover, we can also say that \(x \sqsubseteq f(x)\), being \(x\) the least element of the set of prefixed points of \(f\).
  Hence, \(x\) is a fixed point of \(f\).
  Last, we notice that each fixed point of \(f\) is also prefixed, so \(x\) is also the least fixed point of \(f\).\qedhere
\end{proof}

\begin{thm}[Kleene's Fixed point theorem]\label{thm:cpo-fixed}
  Let \(f\) be a continuous function from an \(\omega\)--cpo \(D\) to itself. %chktex 21
  Then, \(\lbrace f^n(\bot) \rbrace_{n \in \mathbb{N}}\) is an ascending \(\omega\)--chain in \(D\) and %chktex 21
  \begin{equation}
    \fix(f) = \bigsqcup_{n \in \mathbb{N}} f^n(\bot)
  \end{equation}
  is the least fixed point of \(f\).
\end{thm}

\begin{thm}[Knaster--Tarski]\label{thm:knaster-tarski} %chktex 8
  Suppose that \((D, {}\sqsubseteq)\) is a complete lattice, that is, a partial order which has greatest lower bounds of arbitrary subsets.
  Then, for a monotonic function from \(D\) to itself,
  \begin{equation}
    \fix(f) := \bigsqcap\,\,\lbrace d \in D \colon f(d) \sqsubseteq d \rbrace %chktex 21
  \end{equation}
  is the least fixed point of \(f\).
\end{thm}
