\begin{abstract}
  In programming language theory, a proven approach adopted to describe the meaning of programs is given by \emph{denotational semantics}, which interprets each data type with a fitting structure --- usually borrowed from domain theory --- and typed terms as elements of such structure.

  Recursively defined data types, like binary trees and lists, are frequently employed in programming; hence, it is crucial to know how an appropriate domain of interpretation can be constructed.
  This problem amounts to finding a solution to a fixed point equation of domains \(F(D) = D\), where \(F\) is dependent on the recursive type definition at hand.
  % Recursive domain equations hold a prominent role in denotational semantics, as well as in the field of domain theory.
  % %Recursive domain equations play an important role in denotational semantics and type theory.
  % In fact, the recursive specification of datatypes that are commonly encountered in computer science can be achieved by finding the least fixed point of certain functions, thus solving equations of the form \(D = F(D)\), where \(D\) is a domain.
  % %In fact, they allow the definition of recursive program and, more important, recursive specification of datatypes.
  A suitable mathematical setting to describe and solve these equations is given by category theory.
  
  In this work, a canonical construction for solutions of least fixed point equations over categorical domains will be introduced, with a focus on a particular type of categories where morphisms do satisfy certain additional conditions.
  Namely, a brief overview of least fixed point theory for cpos and categories will be presented.
  After that, a dissertation over \(\mathbf{O}\)--categories and solution of recursive equations in such categories will be held.
  Finally, it will be shown how main features of data types of a programming language fit into this framework.
\end{abstract}
